\documentclass[11pt, letterpaper]{practical-letter}

% Sender info
\address{Dept. of Linguistics,\\The University of Texas at Austin}
\phone{+1 020 3870 2020}
\email{gvenkata1994@gmail.com}
\website{venkatasg.net}

% Addressee info
\addressee{Department,\\University,\\ Arrakis CRM 114}
\subject{}

\begin{document}

\senderbox

\opening{To whomever it may concern,}

I am writing to apply for the tenure-track position of Assistant Professor in Computer Science. I am a final-year Ph.D candidate in the Department of Linguistics at The University of Texas at Austin, and intend to graduate by May 2024. I specialize in Computational Linguistics and Natural Language Processing (NLP), with secondary research interests in Computational Social Science (CSS) and pragmatics.

My research is unique in \textbf{using insights from modern NLP techniques and literature to answer questions in human communicative behavior} --- my thesis focuses on how intergroup bias manifests in language online. Throughout my graduate studies, I have regularly published my research in core Computational Linguistics \& NLP venues (the annual conference of the Association of Computational Linguistics (ACL), and the Empirical Methods in NLP (EMNLP) conference), as well as in Computational Linguistics journals (Transactions of the ACL). I plan to continue my interdisciplinary research with a focus on \textbf{diversifying my languages and dialects of study}, as well as concentrating on two phenomena --- \textbf{linguistic generalization} and \textbf{semantic invariance}, towards the ultimate goal of building more robust NLP tools as I lay out in my research statement. 

I have been fortunate to have gained considerable experience in teaching as part of my Ph.D. In addition to serving as the Teaching Assistant on several courses in Computational Linguistics, I was the Principal Instructor for a summer undergraduate class introducing NLP and the computational study of language. My experiences have shaped my teaching philosophy that I outline in my teaching statement, and have instilled in me ambitious goals as an educator. I enjoy working with small class groups where I can closely interact with students and hone their argumentative skills --- essential I believe in \textbf{a field at the intersection of liberal arts and technology like Computational Linguistics}. Further, I believe my experience in industry (as an Applied Scientist at Amazon and a Data Scientist at The New York Times) gives me a unique insight into the skills and knowledge companies seek from applicants, and help me guide and mentor students in a way that aligns with their career goals. I am also excited to teach courses at a wide variety of levels and across the curriculum --- from NLP, semantics, pragmatics and machine learning to an introductory course in computational linguistics.

In addition to research and teaching, I have served as a member of the organizing committee on the Texas Linguistics Society conference for \href{http://tls.ling.utexas.edu/2021/}{2021} \& \href{http://tls.ling.utexas.edu/2022/}{2022}, and the \href{https://sites.utexas.edu/sxsemantics}{South by Semantics Workshop} (2022--ongoing). A guiding principle for both was to \textbf{highlight work done by scholars from underrepresented communities} in Linguistics, Philosophy of Language and NLP, especially early-career researchers, to a broader audience at UT Austin. I believe Linguistics and Computer Science as fields are better served the more people we include from diverse and historically marginalized communities. I hope to continue this mission by uplifting and championing the work of students and colleagues.

I believe that my experience and skills in research, teaching and mentorship make me a good fit for your department and University, and that the supportive environment at the department and university will enable me to hone my skills as an assistant professor. I have attached my CV, research statement, teaching statement, diversity statement, graduate transcript, and letters of recommendation with this application. If you need further information from me, please don't hesitate to reach out. I hope to have the opportunity to talk with you soon, and thank you for considering my application.

\closing{Sincerely,}

\end{document}
 
